The fundamental shape of dynamic experiences is preserved in episodic memory

### 2018-07-12 main.tex
checking in vector field methods

\subsubsection{Naturalistic extensions of classic list-learning memory analyses}
Just like in a traditional `free-recall' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. The recalled events can be treated as ``items'' analogous to words recalled in a list-learning study. In our first set of analyses, we sought to characterize memory performance/dynamics by extending classic analyses originally designed for list-learning experiments to more naturalistic settings.

@@

\subsubsection{The fundamental shape of an experience is preserved in recall}
The analyses described in the previous section are useful in that they allow us to quantify memory dynamics for naturalistic stimuli in a way that is comparable to a vast literature on trial-based list-learning experiments. However, these methods fail to capture the rich temporal structure present in naturalistic stimuli and associated recall of those experiences. Thus, our next set of analyses test whether the ``shape'' of an experience is preserved in memory. We hypothesized that despite substantial individual variability in the precision and amount of content recalled across by participants, the overall shape of the episode would be recapitulated by the majority of participants.

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/4_trajectory.pdf}
\caption{\small \textbf{Video and recall trajectory plots.} A). 2-dimensional embedding of the video events model. The arrows indicate the forward direction of the video events. B). 2-dimensional embedding of the average recall events model.  The colors refer to the most similar video events. The lines connecting the points represent the average probability of transitioning from one event to the other. The opacity of the line is proportional to the transition probability (0 to 1). Arrows indicate the transition with the highest probability for each event.}
\label{fig:trajectory}
\end{figure}

@@

\subsubsection{Measuring the quality of recall}
Representing the video and verbal recall as events in `topic space' also affords us the ability to characterize the quality of recall in a more fine-grained and nuanced way than what was previously possible. To quantify the similarity between the video model and individual recall models, we considered a number of novel metrics.  First, we tested whether each participant's recall model matched the movie model in a general sense. To do this, for each participant we filtered the video model to only include the events that the participant remembered. Then, we computed the root mean squared difference (RMSD) between the video model and the recall model. As an example, if the participant remembered all the events in order (with perfect precision), the expected distance value would be 0. However, if they remembered a subset of events, events our of order or with low precision the expected distance would be greater than 0. To assess significance, we performed a permutation test where we circularly shifted the recall model (10000 times) and recomputed the RMSD. The recall model significantly matched the video model for nine of the participants ($p$<.05; participants: 3-4, 8-13, 17 and the p-value for the rest of the participants was less than .25). Furthermore, the RMSD values were significantly correlated to memory performance across participants (Pearson's $r$(16)=-.57, $p$=.016). Thus, a closer match between the video and recall event models was related to better recall performance.

@@

Our metric for choosing the `best fitting' HMM was to choose the model with the $k$ value that maximized the ratio of the average `within-event' correlation values (i.e. the correlation values for blocks of consecutive timepoints the model identified as one event) and the average `across-event' correlation (i.e. the rest of the correlation values). Additionally, we included a penalty parameter that was proportional to the smoothing of the model that preferred models with smaller $k$ values. We chose $k$ values separately for the video model and for each recall model.  Then, using the best $k$ values, We fit a separate HMM for each topic model. Finally, we averaged over timepoints identified to be in the same event resulting in a $k$ events by 100 topics matrix for the video model and each of the recall models.
%
\subsubsection{Matching recall events to video events}
To figure out which video event each recall event referred to, we correlated the video events model and each recall events model. This resulted in a video events (34) by recall events (8-27) correlation matrix (for each participant) which contains the similarity between each video event and each recall event.  To find the most likely video event that a given recall event referred to, we computed the argmax over the columns of this matrix.  The result was a list of video event indices for each participant. These indices are analogous to the values found in a ``recall matrix'' from a free recall list learning experiment, but represent the recall of particular events (instead of words, for example).
%
\subsubsection{Characterizing memory performance using classic analyses.}

@@

>>>>

\subsubsection{Characterizing naturalistic memory with traditional list-learning analyses}
Just like in a traditional `free-recall' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. The recalled events can be treated as ``items'' analogous to words recalled in a list-learning study. In our first set of analyses, we sought to characterize memory performance/dynamics by extending classic analyses originally designed for list-learning experiments to more naturalistic settings.

@@

 \subsubsection{Measuring the quality of naturalistic memories}
 Representing the video and verbal recall as events in `topic space' also affords us the ability to characterize the quality of recall in a more fine-grained and nuanced way than what was previously possible. To quantify the similarity between the video model and individual recall models, we considered a number of novel metrics.  First, we tested whether each participant's recall model matched the movie model in a general sense. To do this, for each participant we filtered the video model to only include the events that the participant remembered. Then, we computed the root mean squared difference (RMSD) between the video model and the recall model. As an example, if the participant remembered all the events in order (with perfect precision), the expected distance value would be 0. However, if they remembered a subset of events, events our of order or with low precision the expected distance would be greater than 0. To assess significance, we performed a permutation test where we circularly shifted the recall model (10000 times) and recomputed the RMSD. The recall model significantly matched the video model for nine of the participants ($p$<.05; participants: 3-4, 8-13, 17 and the p-value for the rest of the participants was less than .25). Furthermore, the RMSD values were significantly correlated to memory performance across participants (Pearson's $r$(16)=-.57, $p$=.016). Thus, a closer match between the video and recall event models was related to better recall performance.

\subsubsection{The fundamental shape of an experience is preserved in recall}
The analyses described in the previous sections are useful in that they allow us to quantify memory dynamics for naturalistic stimuli in a way that is comparable to a vast literature on trial-based list-learning experiments. However, these methods fail to capture the rich temporal structure present in naturalistic stimuli and associated recall of those experiences. Thus, our next set of analyses test whether the ``shape'' of an experience is preserved in memory. We hypothesized that despite substantial individual variability in the precision and amount of content recalled across by participants, the overall shape of the episode would be recapitulated by the majority of participants.

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/4_trajectory.pdf}
\caption{\small \textbf{Video and recall trajectory plots.} A). 2-dimensional embedding of the video events model. The arrows indicate the forward direction of the video events. B). 2-dimensional embedding of the average recall events model.  The colors refer to the most similar video events. The directional lines connecting the points represent the true video event order. The arrows represent the average direction of all recall event transitions (i.e. a line segment connecting two consecutive recall event vectors) that intersected a box (width=.25) centered on the origin of the arrow.
\label{fig:trajectory}
\end{figure}

@@

@@

Our metric for choosing the `best fitting' HMM was to choose the model with the $k$ value that maximized the ratio of the average `within-event' correlation values (i.e. the correlation values for blocks of consecutive timepoints the model identified as one event) and the average `across-event' correlation (i.e. the rest of the correlation values). Additionally, we included a penalty parameter that was proportional to the smoothing of the model that preferred models with smaller $k$ values. We chose $k$ values separately for the video model and for each recall model.  Then, using the best $k$ values, We fit a separate HMM for each topic model. Finally, we averaged over timepoints identified to be in the same event resulting in a $k$ events by 100 topics matrix for the video model and each of the recall models.
%
\subsubsection{Matching recall events to video events}
To figure out which video event each recall event referred to, we correlated the video events model and each recall events model. This resulted in a video events (34) by recall events (8-27) correlation matrix (for each participant) which contains the similarity between each video event and each recall event.  To find the most likely video event that a given recall event referred to, we computed the argmax over the columns of this matrix.  The result was a list of video event indices for each participant. These indices are analogous to the values found in a ``recall matrix'' from a free recall list learning experiment, but represent the recall of particular events (instead of words, for example).
%
\subsubsection{Characterizing memory performance using traditional approaches.}

@@

\subsubsection{Visualizing the video and recall event models.}

To better understand the temporal structure of the video event model (34 events by 100 topics) and the recall event models (8-27 events by 100 topics), we used a technique called UMAP to reduce the ``topic-space'' from 100 dimensions down to 2 dimensions (REF). UMAP is a nonlinear dimensionality reduction technique which models the manifold of the data with a fuzzy topological structure, and then searches for a (2D) projection of the data that has the closest equivalent fuzzy topological structure. We concatenated (vertically stacked) all event models (video, average recall, and individual recall), and then fit and transformed all of the models at once. This assured that the models were projected into the same space.

\subsubsection{Vector field analysis.}
To quantify (and visualize) the flow of recall from event to event, we performed a vector field analysis.  We tiled the 2D topic space (x, y: -6 to 6 by .25) with an evenly spaced grid. For each grid point, we drew a square around the point (height/width=.5).  Then, we tested whether any line segments (formed by event recall transitions) passed through this area of the topic space.  For example, say that a participant transitioned from recalling event 2 to event 3. These 2 recall events correspond to 2 points in topic space, and connecting them forms a line segment. We collected all line segments that passed through a given section of topic space (collapsing across participants). To plot the average direction of the line segments (i.e. the arrows for each grid point in ~\ref{fig:trajectory}b), we converted each of them to unit vectors and then averaged. For grid points where the direction was consistent (across all participants contributing to that point), the length of the arrow approaches 1, whereas if the direction was random the length of the arrow approaches 0. Lastly, we converted each unit vector to an angle (in radians) by taking the inverse tangent of the x, y components of the vector. To test whether the distribution of angles was significantly non-uniform (i.e. displayed a preferred direction across participants), we performed a Rayleigh test on the angles (REF). Arrows where the Rayleigh test was significant are displayed in color while non-significant tests are displayed in gray with lower opacity.
